test_imputed

### 2015-12-16 finalwriteup/draft.rmd
thesis update

This project will tackle the problem of estimating fixed fees for complex consulting projects. It is therefore imperitive to understand what research has been performed to date and how this project contributes to the body of work. The task of estimating fixed fees can be likened to the terms 'cost estimation' or 'effort estimation' in the literature. The bulk of research to date has been performed with project data in either the construction industry or software development. There is also a body of research on predicting returns on movies in the film industry. The research into cost estimation in each of these three industries is reviewed below, detailing different methods and outcomes. Finally a gap in the literature will be highlighted which this research aims to fill.

@@

Detailed analysis refers to the process of an engineer or builder in the construction team carefully reviewing construction drawings and sum the cost of materials, labour, machinery hire, overheads, profit etc. [@Akintoye2000].  @Akintoye2000's study based in the UK found that this is the most popular method for pricing amongst contractors. The second most popular method was to perform an additional manual comparison with similar projects and make judgments based on personal experience. Note that this does not involve a mathematical model.

@Elfaki2014 discusses how much detailed cost estimates for cosntruction projects can vary from estimator (engineer) to estimator, which contributes to the lack of accuracy in predciting final cost of a project. He argues that an engineer's expertise is not documented or measured in any way and therefore their expertise and therefore estimate is prone to subjectivity. @Shane2009 theorise that final cost can be subject to so many different unpredictable parameters, such as weather, unexpected ground conditions, duration or sub-contractor issues. It is therefore almost impossible to achieve an accurate cost prediction manually [@Elfaki2014].

Groundbreaking methods into calculating construction costs by @Ma2014 mobilise the data from building information modelling, although this method is still technically a detailed analysis of the costs. The idea is to quantify the cost of building a structure directly from a three dimensional model created by the design consultants such as engineers and architects. @Ma2014's first trial was to automate the cost calculation of structure composed of reinforced concrete only. His chief development was to program his algorithm to intelligently establish construction techniques for each element as this is not provided by architectural or engineering models. From the construction technique and material, a cost can be determined for each element. This was successful and Ma aim's to further his work so that more construction information can be intelligently obtained from BIM models. A system like Ma's would definitely accelerate the cost estimation process and reduce human error, however it is still an 'inside view' and could be prone to innaccurate estimates in a similar way to a detailed analysis by a person. The detailed view does not 'step back' and take into account set backs in construction that may have affected similar projects in the past. There is potential for this to be built into the same system but this is still in early stages of development.

Survey information on the latest techniques for cost estimation is limited, however what is available indicates that cost estimation through detailed analysis (inside view) remains the most prevalent method. This is despite construction having a long history of projects that have run over time and over budget using the same cost estimation technique [@Shane2009]. There is even current research on how to automate this way of cost estimation using BIM. The method of detailed sums shall be compared to the results of studies using parametric algorithms and case based reasoning ('outside view') in the following sections.

@@

As the problem of cost estimation in construction projects is well recognised, dozens of studies have modelled the financial performance of these projects using machine learning and statistical models. The aim is to test whether a mathematical comparison of many similar projects, using meaningful descriptive variables, can predict the cost of construction. This section will first review the previous studies and most popular algorithms and finally summarise industry uptake of the methods.

@Elfaki2014 completed a study on cost estimation techniques and research from 2004 to 2014. It was found that artificial neural networks and support vector machines were the most common machine learning techniques used. He highlighted that machine learning techniques deal well with uncertainty, however lack technical justification. In a way, this fulfils the purpose of the 'outside view', which is to step back from the detailed analysis and perform a comparison with similar projects.

@Shin2015 was the first

parametric estimating is not popular with contractors. this could be because:
The reasons include: (a) lack of familiarity with the techniques; (b) the degree of sophistication involved in the techniques is unwarranted for project performance; (c) time plus lack of information and knowledge; (d) doubts whether these techniques are applicable to the construction industry; (e) most construction projects are seldom large enough to warrant the use of these techniques or research into them; (f) they require availability of sound data to ensure con®dence;

Recent developments in cost esti- mating methods and tools that consider risks and variability in cost estimates, such as the use of range estimating and parametric estimating techniques, have not been adopted by contractors.

At the conclusion of the papers, few compare the performance against the original method????

* case based reasoning

### 2.1.2 Previous work on Regression and Neural Networks?

* managers currently go by gut feeling or look up a couple of past examples
* difficult to combine insights from similar jobs/clients, so complex

### 2.1.3 Gap
* no mobilisation of CRM data into sophisticated models in B2B world

## 2.2 Use cases of CRM data

* no idea
* gap - my project obviously

@@

* anova

* linear regression
    * pros

    * cons
        * do these acknowledge their data is not normally distributed?

>>>>

This project will tackle the problem of estimating fixed fees for complex consulting projects. It is therefore imperitive to understand what research has been performed to date and how this project contributes to the body of work. The task of estimating fixed fees can be likened to the terms 'cost estimation' or 'effort estimation' in the literature. The bulk of research to date has been performed with project data in either the construction industry or software development. The research into cost estimation in each of these industries is reviewed below, detailing different methods and outcomes. Finally a gap in the literature will be highlighted which this research aims to fill.

@@

Detailed analysis refers to the process of an engineer or builder in the construction team carefully reviewing construction drawings to sum the cost of materials, labour, machinery hire, overheads, profit etc. [@Akintoye2000].  @Akintoye2000's study based in the UK found that this is the most popular method for pricing amongst contractors. The second most popular method was to perform an additional manual comparison with similar projects and make judgments based on personal experience. Note that this does not involve a mathematical model.

@Elfaki2014 discusses how much detailed cost estimates for cosntruction projects can vary from estimator (engineer) to estimator, which contributes to the lack of accuracy in predciting final cost of a project. He argues that an engineer's expertise is not documented or measured in any way and therefore their expertise and, in turn, their estimate is prone to subjectivity. @Shane2009 theorises that final cost can be subject to so many different unpredictable parameters, such as weather, unexpected ground conditions, duration or sub-contractor issues that it is almost impossible to achieve an accurate cost prediction manually.

Groundbreaking methods into calculating construction costs by @Ma2014 mobilise the data from building information modelling, although this method is still technically a detailed analysis of the costs. The idea is to quantify the cost of building a structure directly from a three dimensional model created by the design consultants (engineers and architects). @Ma2014's first trial was to automate the cost calculation of a reinforced concrete structure. His chief development was to program his algorithm to intelligently establish construction techniques for each element as this is not provided by architectural or engineering models. From the construction technique and material, a cost can be determined for each element. This was successful and Ma aim's to further his work so that more construction information can be intelligently obtained from BIM models. A system like Ma's would definitely accelerate the cost estimation process and reduce human error, however it is still an 'inside view' and could be prone to innaccurate estimates in a similar way to a detailed analysis by a person. The detailed view does not 'step back' and take into account set backs in construction that may have affected similar projects in the past. There is potential for this to be built into the same system but this is still in early stages of development.

Surveyed information on the latest techniques for cost estimation is limited, however what is available indicates that cost estimation through detailed analysis (inside view) remains the most prevalent method. This is despite construction having a long history of projects running over time and over budget using the same cost estimation technique [@Shane2009]. There is even current research on how to automate this way of cost estimation using BIM. The method of detailed sums shall be compared to the results of studies using parametric algorithms and case based reasoning ('outside view') in the following sections.

@@

As the problem of cost estimation in construction projects is well recognised, dozens of studies have built prediction models of these projects using machine learning and statistical models. The aim is to test whether a mathematical comparison of many similar projects, using meaningful descriptive variables, can predict the cost of construction. This section will first review the previous studies and most accurate algorithms and finally summarise industry uptake of the methods.

@Elfaki2014 completed a study on cost estimation techniques and research from 2004 to 2014. It was found that artificial neural networks and support vector machines were the most common machine learning techniques used. He highlighted that these two machine learning techniques deal well with uncertainty, however lack technical justification to the decision maker (known as a black box). Neural networks are also time consuming to train, and must be re-trained and re-tested with each additional piece of data [@Kim2004]. Nevertheless, neural networks and SVM's recieved a significant amount of attention in the 1990's for their ability to accurately predict construction costs with limited detailed information [@Shin2015][@Kim2004].

@Shin2015 pioneered the application of boosted trees to cost estimation in construction projects. This is surprising given the rapid uptake of boosted trees over the past decade. In the study, data from 234 school building construction projects in Korea was used. Boosted trees were compared to neural networks, which are one of the most popular algorithms to apply to the problem. Boosted trees were found to predict costs more slightly accurately than neural networks, but not significantly (p value < 0.5). Even though the two algorithms performed similarly, the output from boosted tree models provide insight into the structure of the model such as variable importance and variable partial dependence plots. This additional insight may mean boosted trees are preferable over neural networks because of their potential to engage decision makers.

A notable amount of literature has studied the predictive power of multiple linear regression in the construction cost estimation problem. Often, regression is the only model assessed, without comparison to other methods such as neural networks, which first started appearing in literature in the 1990's [@Kim2004]. However, even post-1990, many studies focused on regression only. This may be because the technique is straightforward, easy to use and widely available in statistical packages [@Chan2005]. Out of those that have compared regression to neural networks, the studies have shown that neural networks outperform regression, however other studies show they are approximately equal [@Kim2004][@Attalla2003]. @Dissanayaka1999 found that neural networks outperformed multiple linear regression models, however regression models could be performed first to determine variable importance and condense the number of variables. It can be argued that boosted trees could perform as well as neural networks while providing additional insights (even more so than regression models) such as variable importance and partial dependence of variables against cost.

Despite neural networks typically outperforming regression models, it is intersting to note some of the outcomes of regression studies in cost estimation. Different studies tend to produce a wide range of variables that contribute to different degrees in the models. This is likey due to the nature of construction projects which are multidisciplinary and involve a range of different parties such as the clients, consultants, contractors and suppliers, each of which can have a range of characteristics and degrees of input [@Chan2005]. @Chan2005's study on 87 building projects in Singapore included special, complex projects. Therefore, variables such as contractor's specialised skills, who the client was (public vs private), the client level of experience, and the contractor's financial management highly influenced final cost of the project. Other studies found variables such as project complexity, duration, team experience, information availablility, site requirements, and labour climate to contribute most [@Akintoye2000][@Trost2003][@Pinto1988]. As these studies obtained information from questionaires, the available variables also largely depended on the questions in the questionaires, and therefore each study produces a range of different important variables. Despite the differences in qualitative outcomes, the studies all generally report their mean absolute error or a similar metric from the tested models. This is encouraging, however there is a gap in the literature in how to translate the improved errors from the model into an argument for industry to uptake the model as an improvement on their bottom line.

Surveys into construction industry uptake of these algorithmic models are sparse, however a survey by @Akintoye2000 showed that parametric estimating methods have not been adopted by contractors, despite the research into this area over 20 years prior. The majority of contractors use experience based methods. @Akintoye2000 hypothesises that this could be due to a lack of familiarity and knowledge about the techniques, doubts whether these techniques are applicable to the construction industry, and the availability of sound data to ensure confidence. This presents a clear gap in the literature on how to analyse the models in terms of the goals and limitations that a contractor may experience.

The literature on cost estimation for building projects do not break down the cost estimation task into smaller components, even though a construction project is completed by dozens of consultants and sub-contractors. Many of the consultants, such as Engineers and Architects, produce their own cost estimates for their effort that contributes to the project. Their contracts are often fixed price, with minimal room for movement, which means they wear the risk of over-spending to meet fixed cost contracts [@Harris1999]. Furthermore, the existing studies collect data from a wide range of projects that have certain characteristics in common, such as country, or building type and information is obtained through surveys. This leaves a gap in research exploring a single contractor's history of projects, which would be available in their company database.

**case based reasoning**

Case based reasoning (CBR) is a method of estimating by using the results of similar cases. The idea is that in the case of a new project, similar projects are chosen, before generalisations about the data are made. This contrasts to algorithmic models that are built from the entire set of data [@Elfaki2014]. Research into the application of case based reasoning to construction cost estimation began in the 1980's [@Kim2004]. The general procedure involves:

1. Storing a collection of projects with key variable values
2. Once a new project or case arrives, similar cases are retrieved. This can be achieved either algorithmically with a distance function applied to the variables, or manually by users reviewing past cases. The nearest neighbour algorithm has been used to find the distance between cases in previous studies. It is programmed to calculate the euclidean distance in n-dimensional space between cases, where each variable is a dimension [@Kumar2007].
3. The cost of the new project is estimated by extrapolating characteristics of the similar cases to the new case. This can also be done either algorithmically or manually by an experienced decision maker.

A study by @Kim2004 compared CBR to neural networks and regression in the construction cost estimation problem. For CBR, the ESTEEM software package was used which algorithmically calculated the similarity of variables, weighted the variables using gradient descent, and deduced the cost using the most similar cases (the specific method for this last step was not explained). The discussion compared mean absolute error rates of the three methods and showed that CBR was more accurate than regression, but less accurate than neural networks in cost estimation. Despite neural networks outperforming CBR, CBR maintained significant advantages over neural networks which have been referenced in other studies as well.

Advantages of CBR include algorithmic efficiency as well as user engagement. In @Kim2004's study, it was noted that CBR models were simple to update with new data in comparison to neural networks, which must be slowly re-trained and re-tuned. New cases also did not need to have every variable complete which can occur often in real, messy data. Neural networks on the other hand require complete data. Another significant advantage of CBR is the ability for users to review chosen similar cases and make sense of the predicted cost, as opposed to neural networks which are a black box [@Kim2004][@Elfaki2014][@Kumar2007]. A disadvantage of CBR is the accuracy can be highly dependent on the number of selected cases [@Elfaki2014]. Overall, CBR may provide prediction accuracies between regression and neural networks but its results can be justified by the user - a valuable asset.

CBR can be described as a systematic method of expert judgment, where the decision maker manually compares similar projects from his or her experience [@Shepperd1996]. CBR that is performed manually through personal experience, without the assistance of algorithms, was the second most popular method for construction estimation in @Akintoye2000's survey in the UK. This indicates it is quite an intuitive method for decision maker's to cross check their detailed cost estimate, although an algorithmic version of CBR has not been taken up by industry.

**summary**

In summary, research into cost estimation in the construction industry has been undertaken using a variety of methods ranging from detailed manual analysis, algorithmic models, and case based reasoning. Neural networks generally outperformed all other systematic methods, such as regression and case based reasoning, but had a significant disadvantage of being a black box. Alternative methods such as boosted trees have the potential to perform as well as neural networks and provide insight into the structure of the algorithm. This is true for CBR as well, although it has not been shown to predict as accurately as neural networks. There exists a gap in the research in ways to intelligently combine CBR with machine learning methods that could predict well, provide insight to the structure, and engage the user in reviewing similar projects. Also, as stated previously, there exists a gap in the application of cost estimating in the construction industry to smaller components of the project, such as consulting and contracting companies who each face a similar problem in estimating their fixed fee. These companies could mobilise information in their internal project databases, as opposed to available literature which surveys dozens of companies involved in selected construction projects.

### 2.1.1 Effort Estimation in the IT Industry

Like the construction industry, the software industry faces the challenging task of estimating the cost of a project before full engagement with their client. However, unlike construction projects, the major cost associated with the project is *effort* as opposed to the cost of building materials. A parallel can be drawn between software project effort estimation research and consultants effort in the construction industry, which was a gap in the research of the previous section.

Effort estimation is indeed a problem, and was demonstrated by a review of 6 surveys of IT projects between 1984 and 1994 by @Molokken2003. It revealed that 60-80% of IT projects encounter effort and/or schedule overruns, where the average overrun was by 30 to 40%. A survey of software managers asked whether they saw effort estimation as a problem, and 91% responded 'yes', while only 9% answered 'no' [@Moores1992]. The most popular response for the supposed reason for cost overruns was over-optimistic estimates (51%), which falls squarely within the theory of optimism bias, or the 'inside view' [@Flyvbjerg2011]. Given this widespread problem in the industry, plenty of research has been dedicated to the problem of effort prediction, with methods similar to the construction industry literature: detailsed analysis, algorithmic modelling, and case based reasoning.

Similar to the construction industry, expert judgment or detailed analysis, is the most widely practised method for effort estimation [@Shepperd1996][@Molokken2003]. This is despite many years of research being dedicated to developing algorithmic models and industry tools that, in the research context, outperform expert judgment. A survey by @Moores1992 found that software tools assisting a software project were in significant use, but only the project planning tools, not the cost estimating tools. This suggests a mental link for project managers to price jobs based on *details* of the software project as opposed to taking an outside view. Another study by @Heemstra1992 found that there was no evidence estimation accuracy improved when estimation tools were used. @Bergeron1992 similarly found jobs that used algorithmic models were actually associated with less accurate estimtes. However, this trend may be coincidental due to a lack of cases where estimation tools were used [@Molokken2003]. The scarcity of real project evidence that algorithmic estimation tools improve estimations and an unintuitive mental jump could be reasons contributing to detailed expert analysis remaining the most widespread technique for effort estimation.

**Algorithmic methods**

Algorithmic methods to predict effort for IT project costs has been a popular subject of research. Regression models have been the most substantially tested followed by neural networks [@Finnie1997]. This section will assess the details and performance of the algorithmic models, important variables, applications and finally gaps in the literature.

Multiple studies have shown that neural networks definitively outperform regression models in effort estimation, although regression is the most popular method in the literature [@Finnie1997][@Pai2013][@Matson1993]. Interestingly, several studies have found that even if 15 or so variables are included, often only one variable contributes significantly to the models accuracy: size [@Shepperd1996][@Finnie1997][@Pai2013]. Size can refer to the expected number of lines of code or function point. Function point refers to the amount of business functionality expected from an information system product. This is easier to guess correctly at the beginning of a project, and therefore leads to more accurate predictions [@Finnie1997]. One study used a single variable approach (size) with a linear coefficient and an exponential coefficient as follows:

$effort = \alpha*size^{\beta}$
$\alpha = productivity coefficient$
$\beta = economies of scale coefficient$
$size = estimated lines of code$

This model was compared to case based reasoning, and case based reasoning outperformed it [@Shepperd1996].

There are disadvantages to algorithmic methods, similar to the construction industry. First, there are often not enough cases to create a good model in the software industry, particularly if the cases must be from within the company [@Finnie1997][@Pai2013]. A study by @Mendes2004 demonstrated using 67 web projects that cross-company models were significantly less accurate than a within company model, so it is in the company's interest to create an in-house model. This means that effort estimation tools should use in-house data and must be adapted by statistical experts to each company [@Shepperd1996].

Algorithmic estimation tools have not been successfully adopted by industry. This may be because no model has proved to be outstandingly successful at consistently predicting required effort [@Finnie1997]. Also, neural networks (the most accurate algorithm) lack the ability to capably explain its results and instill trust in decision makers [@Finnie1997].

The literature heavily focuses on effort estimation for the software industry. This leaves a gap in studies for effort estimation for consulting companies in other industries such as the construction industry. Construction projects do not have a simple 'project size' or function point variable and would need to utilise other variables available.

**Case Based Reasoning**

In the IT industry, applications of case based reasoning to effort estimation have been researched in a similar manner to the construction industry. The procedure is described as a algorithmicallyfinding similar projects to a new case using measurements of euclidean distance in n-dimensional space, where each dimension corresponds to a variable [@Shepperd1996]. The similar cases are then used to predict effort for the new case, and in @Shepperd1996's study, linear regression was used for this final step.

In contrast to the construction industry, the literature in software effort estimation found CBR to perform approximately equally to algorithmic methods such as neural networks [@Shepperd1996][@Finnie1997]. It was also highlighted that CBR is intuitively similar to how an expert thinks about pricing projects and provides insight into the prediction that neural networks do not [@Finnie1997]. Research into CBR in the IT industry found similar results to the construction industry except that CBR performed as well as algorithmic methods.

**Summary**

Research into effort estimation in the software industry has followed a similar path to the cost prediction of construction projects. Expert judgment via detailed analysis is still the predominant method for effort estimation. Case based reasoning and algorithmic methods such as neural networks have found some success in effort estimation (at similar levels) but have not been successfully adopted by industry. It was noted that case based reasoning resonated more with decision makers because its effort prediction could be justified by the model. This visibility of the model prediction structure is very important for future work if the results are to be translated to industry. Effort estimation remains a problem, however agile methods of project management and effort management are changing the contractual approach of consulting in the software industry [@Badenfelt2011]. This innovative method keeps delivery times short (and therefore predictable effort) and defines projects as something with a fixed cost and time. This means if a change occurs, other less important elements must be discarded maintain the fixed amount. Software development is a relatively new field in comparison to the construction industry and is able to adopt this new model relatively quickly across the industry. However, effort estimation for consulting companies in the construction industry are faced with a long history of tradition that is much harder to tackle. This leaves a gap in research in algorithimic models that predict effort for consulting companies in the construction industry. From previous literature, the models should allow decision makers to interrogate the structure of the prediction and be based on internal company data.

## 2.2 Statistical and machine learning methods

### 2.2.3 Introduction

### 2.2.4 Linear Regression/logistic regression

### 2.2.5 Naiive Bayes

### 2.2.6 Ensemble Trees

### 2.2.7 Bayesian Networks

### 2.2.7 SVM's and Neural Networks

### 2.2.8 Summary

@@

### 2015-12-16 rf_log_naiive_auc.rmd
thesis update

#partial dependence plots
fit.boost$var.names
plot.gbm(fit.boost, 9, type="response")
biz<- plot.gbm(fit.boost, 7, type="response")

@@

pred.boost <- predict(fit.boost, test, n.tree = 10000, type = "response")
pred.p <- roc(test$f.rpdol, pred.boost)

>>>>

#partial dependence plots
fit.boost$var.names
plot.gbm(fit.boost, 3, type="response")
biz<- plot.gbm(fit.boost, 7, type="response")

@@

pred.boost <- predict(fit.boost, test, n.tree = 10000, type = "response")
pred.p <- roc(test$b.rpdol, pred.boost)
